---
title: "CPT analysis"
author: "Anthony Sonrel"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: yeti
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Some abbreviations: 

- map = mean arterial pressure

- t0 = first baseline measure

- t1 = second baseline measure

- t2 = measure during CPT

- hr = heart rate (secondary outcome)

## Prerequisites to run the report

Data, with and without labels, should be downloaded from RedCap and placed in a `data` folder at the same level as this Rmd file. 


```{r}
list.files("data")

suppressPackageStartupMessages({
  library(ggplot2)
  library(reshape2)
  library(patchwork)
  library(car)
})
```


## Load data

```{r}
dat_labels <- read.csv("data/EPI301GroupII_DATA_LABELS_2021-11-03_0935.csv")
dat <- read.csv("data/EPI301GroupII_DATA_2021-11-03_0938.csv")
```

Add the room number : manual mapping from the schedule of the day where we decided that participants from left and right columns were assigned to different rooms. This will be used latter on to check if we have a 'room effect'. 

```{r}
dat$room <- c("left", "left", "right", "right", "left", "right", "right", "left", "right", "right", "right", "right", "left", "left", "left", "left", "left", "right", "left", "left", "right", "left")
```


Remove any incomplete data at any of the timepoints. This will remove the participants who didn't complete the questionaire, who removed their hand and the one with machine error. 

```{r}

to_keep <- !is.na(dat$map_t2)
dat <- dat[to_keep, ]
dat_labels <- dat_labels[to_keep, ]
table(to_keep)

```



Let's set the baseline as the minimum map between t0 and t1 for each participant, for MAP and HR. 

```{r}

dat$map_baseline <- apply(dat, 1, function(x){
  as.numeric(min(x["map_t0"], x["map_t1"]))
})

dat$hr_baseline <- apply(dat, 1, function(x){
  as.numeric(min(x["hr_t0"], x["hr_t1"]))
})

```

Replace some cryptic values: 

```{r}
dat$sex <- ifelse(dat$sex == 2, "F", "M")
dat$randomization_group <- ifelse(dat$randomization_group == 1, "Control", "Distraction")
```

Rename treatments and see how people were assigned: 

```{r}
table(dat$randomization_group)
table(dat$randomization_group, dat$sex)
whichA <- which(dat$randomization_group == "Control")
whichB <- which(dat$randomization_group == "Distraction")
```

BMI calculation to assess its effect latter on: 

```{r}
dat$BMI <- dat$weight/((dat$height/100)^2) 
summary(dat$BMI)
```

Deltas of MAP and HR: 

```{r}
dat$map_delta <- dat$map_t2 - dat$map_baseline
dat$hr_delta <- dat$hr_t2 - dat$hr_baseline
```

Remove the 2 participants which are >40y.o. **We'll keep the data of these 2 patients for a sensitivity analysis**, ie assess if the inclusion of these 2 patients had an effect on the analysis. 

```{r}

dat$age <- 2021 - as.numeric(gsub("-.*", "", dat$dateofbirth))
to_keep <- dat$age < 40
dat40yo <- dat[!to_keep, ]
dat <- dat[to_keep, ]
```


## Overview of data

```{r}
str(dat)
```

```{r}
summary(dat)
```


### MAP


Distribution of MAP across timepoints

```{r}
boxplot(dat$map_t0, dat$map_t1, dat$map_baseline,  dat$map_t2, 
        names = c("T0", "T1", "min(t0-t1)", "t2"),
        main = "MAP:  all participants")
```

Same, divided  by intervention 


```{r, warning=FALSE}
dat_melt <- melt(dat, id.vars = c("randomization_group" ), measure.vars = c("map_baseline", "map_t2"))
dat_melt$variable <- factor(ifelse(dat_melt$variable == "map_baseline", "baseline", "CPT"))

ggplot(dat_melt, aes(x = variable, y = value, fill = randomization_group))+
  geom_boxplot() + 
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.8))+
  xlab("Timepoint") + ylab("MAP [mmHg]") +
  theme(axis.text.x = element_text(18)) +  
  labs(fill='Randomization group') 

```

Which can also be observed on a scatterplot, more relevant for the linear models that we will be testing (see `data analysis`). 

```{r}
## 1 column map baseline, 1 column map cpt, 1 column randomization group

ggplot(dat, aes(x = map_baseline, y = map_t2, 
                color = randomization_group))+
  geom_point(size = 3) + xlab("MAP baseline [mmHg]") + ylab("MAP CPT [mmHg]")+ 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE) +  
  labs(color='Randomization group') +
 geom_abline(aes(slope=1, intercept=0), size=1) 
```


Secondary outcome: **heart rate** 

```{r}

dat_melt <- melt(dat, id.vars = c("randomization_group" ), measure.vars = c("hr_baseline", "hr_t2"))
dat_melt$variable <- factor(ifelse(dat_melt$variable == "hr_baseline", "baseline", "CPT"))

ggplot(dat_melt, aes(x = variable, y = value, fill = randomization_group))+
  geom_boxplot() + 
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.8))+
  xlab("Timepoint") + ylab("HR [BPM]")  +  
  labs(fill='Randomization group') 

```


Secondary outcome: **percieved pain** 


```{r}
boxplot(dat$pain[whichA], dat$pain[whichB], 
        names = c("Control", "Distraction"),
        main = "Pain:  all participants")
```


### Possible confounders

Let's check the values among some possible confounders such as sex, assigned room, BMI, ...

```{r}
## Sex
dat_melt <- melt(dat, id.vars = c("sex" ), measure.vars = c("map_baseline", "map_t2"))
dat_melt$sex <- factor(ifelse(dat_melt$sex == "F", "female", "male"))
dat_melt$variable <- factor(ifelse(dat_melt$variable == "map_baseline", "baseline", "CPT"))

ggplot(dat_melt, aes(x = variable, y = value, fill = sex))+
  geom_boxplot() + 
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.8))+
  xlab("Timepoint") + ylab("MAP [mmHg]")


```

As expected, we have a difference between males and females, which justifies our stratification strategy. 

```{r}
## Room assignment
dat_melt <- melt(dat, id.vars = c("room" ), measure.vars = c("map_baseline", "map_t2"))
dat_melt$variable <- factor(ifelse(dat_melt$variable == "map_baseline", "baseline", "CPT"))

ggplot(dat_melt, aes(x = variable, y = value, fill = room))+
  geom_boxplot() + xlab("Timepoint") + ylab("MAP")

```

No observable difference between the assigned room so we can rule out bias due to assigned group (procedure, blanket, ice bucket, etc.).

```{r, fig.width=8}
## BMI
dat_melt <- melt(dat, id.vars = c("BMI" ), measure.vars = c("map_baseline", "map_t2"))
dat_melt2 <- melt(dat, id.vars = c("BMI" ), measure.vars = c("sex"))

ggplot(dat_melt, aes(x = BMI, y = value, color = variable, shape = variable))+
  geom_point() + xlab("BMI") + ylab("MAP")+ 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)+
ggplot(dat_melt2, aes(x = value, y = BMI))+
  geom_boxplot() + xlab("Sex") + ylab("BMI")

```

It seems that BMI has an influence on MAP but since it is correlated to sex and that we stratified the data for sex, we can assume that BMI shouldn't affect the results. 

### Deltas

Let's visualize the deltas, as the difference between the second measurement and the baseline. 

```{r, fig.width=8}

par(mfrow = c(1,2))
boxplot(dat$map_delta[whichA], dat$map_delta[whichB], 
        main = "MAP CPT - MAP baseline", 
        names = c("Intervention A", "Intervention B"))
boxplot(dat$hr_delta[whichA], dat$hr_delta[whichB], 
        main = "HR CPT - HR baseline", 
        names = c("Intervention A", "Intervention B"))

```

```{r}
dat_melt <- melt(dat, id.vars = c("randomization_group" ), measure.vars = c("map_delta"))
ggplot(dat_melt, aes(x = variable, y = value, fill = randomization_group))+
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.75))+
   ylab("MAP change [mmHg]") 

```


```{r}
dat_melt <- melt(dat, id.vars = c("randomization_group" ), measure.vars = c("hr_delta"))
ggplot(dat_melt, aes(x = variable, y = value, fill = randomization_group))+
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.75))+
   ylab("HR change [BPM]") 

```

```{r}
dat_melt <- melt(dat, id.vars = c("randomization_group" ), measure.vars = c("pain"))
ggplot(dat_melt, aes(x = variable, y = value, fill = randomization_group))+
  geom_boxplot() +
  geom_dotplot(binaxis='y', stackdir='center',dotsize = 0.8,
                 position=position_dodge(0.75))+
   ylab("Pain level") + ylim(c(0,10))

```

## Data analysis

### Primary outcome

Let's perform a **two-sample T-test on deltas** as we have 2 independent samples and we can assume that the deltas, derived from MAP, are normally distributed. 

```{r}

t.test(map_delta ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```

--> **NOT significant**


Some observations: 

- the sample size is very small and the boxplots show that the standard deviation is quite high. 

- Intervention A was expected to lead to a higher difference, not the other way around. 

### Other tests

I'm testing here other ways of analyzing the data. **I still recommend to report the T-test and its statistics as it's the most straightforward test. To be exhaustive, we can however report that we also compared it with the following tests**. 

ANCOVA, assessing the  additional effects of `randomization_group` and `sex` to evaluate if `sex` was correctly stratified:  

```{r}
ancova_model <- aov(map_delta ~ randomization_group + sex + room, data = dat)
Anova(ancova_model)
```


No effect for `randomization_group`, after having corrected for `sex.` Given that `sex` is not significant, we can remove it from the model. 

```{r}
ancova_model <- aov(map_delta ~ as.factor(randomization_group), data = dat)
summary(ancova_model)
## Confidence interval
confint(ancova_model)
```

Which yields exactly the same values (look at p.value and CI) as for the T-test. 

### Secondary outcomes

We are also interested in the difference of **heart rate** and **pain** felt during the CPT. 

```{r}
## HR
t.test(hr_delta ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```


--> **NOT significant**

```{r}
## Pain
t.test(pain ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```

--> **NOT significant**

### Sensitivity analysis

Two participants which didn't meet the inclusion criteria where tested in our experiment. We can perform a sensitivity analysis to assess whether the inclusion of these two participants had an effect on the final outcome: 

```{r}
## reinclude the 2 particpants
dat <- rbind(dat, dat40yo)
```

```{r}
## MAP
t.test(map_delta ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```

```{r}
## HR
t.test(hr_delta ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```

```{r}
## Pain
t.test(pain ~ randomization_group, data = dat, 
       alternative = "two.sided", var.equal = TRUE)

```

--> The statistics for the 3 tests using the 2 older participants is similar to the tests including them. **We can then be pretty confident that the inclusion/ exclusion of these participants don't change our conclusions. **

## Session info 

```{r}
sessionInfo()
```
